Effects of Solid State Drives on Database Performance

### 2017-04-11 term_paper/paper.tex
Lots more paragraphs.

% Document starts
\begin{document}
% Title portion. Note the short title for running heads 
\title[Short Title]{Effects of Solid State Drives on Database Performance}  
\author{Philip M. Westrich}
\affiliation{
  \institution{Tennessee Technological University}
  \streetaddress{1 William L. Jones Drive}
  \city{Cookeville} \state{TN} \postcode{38501} \country{USA}
}

@@

Several of these grids, usually one to four, are placed together into what is known as a die. Each die can read somewhere 
around 400 MB/sec, while they can only write at around 20MB/sec. This is due to the much more complicated write method 
they employ. However, they have much lower latencies for these operations, typically measured in microseconds, and in 
order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. \cite{Cornwell2012, Micheloni2013}

@@

The controller's job is twofold. First, it must translate the operations requested by the host machine into operations 
that be performed on the NAND flash. Second, it must do this in a manner such that wear on the device is minimized and 
even, and the time needed to perform the requested operations is minimized.

To carry out these tasks, SSDs have a small computer, known as a microcontroller. The microcontroller is also paired with 
specialized hardware to speed up its computations.

@@

>>>>

% Document starts
\begin{document}
% Title portion. Note the short title for running heads 
\title[Effects of SSDs on Databases]{Effects of Solid State Drives on Database Performance}  
\author{Philip M. Westrich}
\affiliation{
  \institution{Tennessee Technological University}
  \streetaddress{1 William L. Jones Drive}
  \city{Cookeville} \state{TN} \postcode{38501} \country{USA}
}

@@

Several of these grids, usually one to four, are placed together into what is known as a die. Each die can read somewhere 
around 400 MB/sec, while they can only write at around 20MB/sec. This is due to the much more complicated write method 
they employ. However, they have a much lower latency for these operations, typically measured in microseconds, and in 
order to meet capacity and speed requirements, multiple dies are packaged inside one SSD. The controller will often 
read and write to multiple dies at once in order to reach performance goals. \cite{Cornwell2012, Micheloni2013}

@@

The controller has many jobs. First, it must present to the host device the same interface HDDs do. Second, it must 
translate the commands issued to it to operations on the NAND flash. Third, it must wear the NAND in the drive evenly 
to prevent premature failure. Fourth, it must deal with the higher error rate that the NAND flash has over HDDs. 
\cite{Cornwell2012, Micheloni2013, MatejFucek2014}

To carry out these tasks, SSDs have a small computer, known as a microcontroller, embedded within them. The microcontroller 
is also paired with specialized hardware to speed up some of its computations, as well as some amount of DRAM to cache and 
buffer events. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

The controller must present to the host device the same interface other block devices do. However, the NAND flash is not 
accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer, or FTL. 
The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

Because cells must be erased before they can be written to, new data is always written to known empty pages, and the old 
pages are marked as garbage in the FTL. Therefore, the actual locations of data may move across the drive as it is written 
to, unlike HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from 
the logical blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes 
a block has experienced as well as a list of bad blocks that have either started out defective from the factory, or have 
worn out over time. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

When the controller is idle for some period of time, it will start performing garbage collection on the now invalid pages 
in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled up 
and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
\cite{Cornwell2012, Micheloni2013, MatejFucek2014}

NAND flash has a much higher error rate than the magnetic platters of HDDs. Therefore, all SSDs have some amount of 
hardware-assisted error correction. Higher quality drives have more dedicated hardware, while cheaper ones may share 
hardware for a larger number of pages. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

@@

For years, systems have assumed that HDDs are the form of persistent storage used, and are optimized for it. These 
optimizations often are detrimental to SSD performance. In this section, we explain these assumptions and why they 
may lead to suboptimal performance.

\subsection{Physical differences}

HDDs have multiple moving parts: the platters that spin at some speed, and the heads that move across them that read the 
data from the platters. It takes time to move these parts through space. Often, to conserve power and lifespan, the drive 
stops rotating its platters. When at a complete stop, it can take between four to ten seconds to bring the platters back 
up to speed. This may not be an issue in consumer-grade applications, but it can be a matter of life and death in critical
real-time systems. \cite{MatejFucek2014} The head of the drive must also move to the correct location to read from the 
requested sector. \cite{Cornwell2012}

Solid state drives have zero moving parts, hence their name. They do not require any spin up time, and their seek time is 
both uniform and much, much smaller. An additional benefit is that they consume much less energy. 
\cite{Cornwell2012, Micheloni2013, MatejFucek2014}

\subsection{Differences in access patterns}

Current systems tend to assume that they will be placing their data on to HDDs, and tend to optimize for that; they try 
to make up for the shortcomings of the physical media. They assume the location of the sectors requested are always in 
the same location, and that lower addresses are towards the faster, outside of the drive, and higher addresses are towards 
the inside. \cite{Cornwell2012}

Because of the seek time added by moving the head across the drive and the assumption of block locations, filesystem drivers 
will perform complicated math to determine how it can move the head the minimum distance for a sequence of reads and writes. 
They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since HDDs are much 
better at large sequential operations. \cite{Cornwell2012}

In order to minimize the number of requests made to the drive, filesystem drivers will typically not inform the drive that 
blocks are freed; they simply mark in their own data structures that the blocks are not needed, and the data remains on 
the disk until something comes to overwrite it. \cite{Cornwell2012}

SSDs break many of these assumptions. Because the FTL takes the liberty of shuffling data around on the disk as it sees 
fit, I/O coalescing and other location assumptions are meaningless, and the time and effort put in to them are wasted. 
Not informing the drive of freed blocks harms the performance of SSDs, as they rely on having as many free pages as 
possible. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

\subsection{What this means for databases}

With the direction that computing is headed, data is invaluable. The databases in which this data is stored are completely 
dependent on the storage medium in which they are backed.
